Supporting Ecological Decision Making Using Feature-Selection and Variable Importance

### 2020-03-16 docs/00-manuscripts/ieee/ieee-manuscript.tex
add conclusion

\noindent We want to stress the differences between 'dimension reduction' and 'high-dimensionality'.
The former refers to the general idea of reducing features from a dataset\cite{vandermaaten2007}.
In modeling this means finding the best subset of covariates which provide the most predictive power to the model or extracting the main components of the features using a \ac{PCA}.
'High-dimensionality' in contrast is a dataset attribute and applies when \(p > n\), where \(p\) is the number of covariates and \(n\) the number of observations\cite{hastie2001}.
There is no absolute value at which the term applies, both for \(n\) and \(p\).
Hence for this study the word ‘high-dimensionality‘ only applies to the
experiments using the ‘NRI‘ feature set (> 7000 features for 1808 observations).

@@

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
\noindent Overall, the response variable 'defoliation at trees' could be modeled with an error of 28 \%.
SVM showed no differences in RMSE across feature sets whereas other learners like RF differed up to seven percentage points (HR-NRI vs. VI) (\autoref{fig:perf-result}).
Ridge and Lasso faced major issues up to the point of not achieving an error below 50 for datasets HR-NRI-VI and NRI-VI.
SVM showed the best performance of all learners with an absolute difference up around three percentage points to the next model (RF) (\autoref{tab:best-learner-perf}).

@@

% learner
% - SVM Car is the overall winner
% - Ridge-CV hit and miss (hit = NRI, miss = VI)
% - all MBO penalized methods have no difference across tasks
SVM combined with the 'Carscore' filter achieved the best performance (RMSE of 27.98) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with Ridge penalty (L2) showed a high variance when comparing results across tasks: In two out of six tasks (all including VI variables) the error was out of bounds (HR-NRI-VI, HR-VI, VI), for HR-NRI and HR the error was around 40 RMSE, and for task NRI the Ridge learner achieved almost the same score as the best performing model (32.9 RMSE) (\autoref{tab:perf-top-10}).
%All penalized methods tuned using MBO showed exactly the same error across all tasks.

@@

@@

\subsection{Permutation-based variable importance}

@@

% FIXME: explain in methods why only 2 datasets are shown
The most important features of each dataset showed an average decrease in RMSE of 1.57 (HR, B69) and 1.79 (VI, Vogelmann2) (\autoref{fig:fi-permut-vi-hr}).
For both datasets most features among the ten most important ones cluster around a wavelength range of 700 nm - 750 nm.
For feature set HR four features in the infrared region (920 nm  - 1000 nm) were identified to be important by causing a mean RMSE decrease of around one percentage point.
Overall, most features show a small importance only with average decreases in RMSE below 0.5.

@@

\noindent The decision to use a buffer of 2 m to generate the index value for each observation is questionable.
When using no buffer at all, the possibility is high that a pixel value gets assigned to a tree which does not spatially match with the actual recorded hyperspectral reflectance due to the geometric offset of 1 m.

Using a buffer of more than 2 meters increases the probability of including information from other trees into the pixel value, blurring the actual value of the tree observation.
In our view using a buffer of 2 m is a good compromise between both worlds.

Another critical point is that the exact number of contributing pixels to the final index value of an observation cannot be determined as it depends on the location of the tree within the pixel grid.
As the buffer is circular, determining the total number of contributing pixels depends on the exact location of a tree within that pixel as this sets the base how many pixels in each direction will be touched by the buffer.
If a tree observation is located at the border of the plot, some directions of the buffer will contain no values and the subsequent index value will be calculated using less pixels than if the tree observation is located in the middle of the plot.

The magnitude of bias introduced by these facts cannot be quantified.
However, it has to be considered when making interpretations about the outcome of this study.

\subsection{Performance vs.\ plot characteristics}
% TODO: Really needed?

\subsection{Predictive Performance}

\subsubsection{Algorithm differences}

\subsubsection{Feature set differences}

@@

% - Ensemble FS show no advantage to simple methods -> maybe to less features, features too correlated? check diff between effect on NRI and VI
% - Effect varies across learners -> suggest to use FS even on feature sets with p < 100?
% - When calculating multiple filters, consider to only use ensemble filter?
% - PCA shows similar performance but probably requires tuning of main components -> check runtime!

@@

\subsection{Comparison to other studies}
% NOTE: no environmental studies use filter methods, some use FFS
% NOTE: only few studies analyze defoliation

\noindent Most other studies analyzing defoliation operated on the plot rather than the tree level.
This originated simply due to spatial resolution of the satellite products that served as the input data\cite{townsend2012, debeurs2008, rengarajan2016}.

Studies focusing on tree-level defoliation used ground-level methods such as \ac{ALS} or \ac{LiDAR}\cite{meng2018, kalin2019}.
\~cite{meng2018} used \ac{OLS} regression methods while\cite{kalin2019} retrieved information out of ground-level RGB photos using \ac{CNN}.
Both study designs are substantially different compared to the setup of this work.
In addition, no spatial \ac{CV} or \ac{FS} was used.
\~cite{goodbody2018} used a \ac{PLS} model with high-resolution \ac{DAP} to predict cumulative defoliation caused by the spruce budworm.
Study results indicated that spectral metrics were found to be most helpful for the model.
Incorporating such metrics (both spectral and structural) could be a possible enhancement for future works.

\~cite{shendryk2016, ludwig2019} are studies which are more similar in their methodology but focus on a different response variable.
\~cite{shendryk2016} used machine-learning models with \ac{ALS} data to study dieback of trees for eucalyptus forests.
A grid-search was used for hyperparameter tuning and a \ac{FFS} for variable selection.
\~cite{ludwig2019} analyzed woody cover in South Africa using spatial \ac{CV} and a relatively novel spatial \ac{FS} approach\cite{meyer2018} on a Random Forest classifier.

In summary, we could not find studies using filter methods for \ac{FS} or \ac{NRI} indices in their work with a relation to forest health.
Most studies used only one algorithm (usually Random Forest) without strong arguments why this particular one has been selected.
In our view the number of possible derivable input features is often higher than the actual number of variables used in the end, missing out on potentially helpful information from the raw data.
These findings underline the importance of this work in the field of forest health modeling.

\section{Outlook and conclusion}

>>>>

\noindent We want to stress the differences between 'dimension reduction' and 'high-dimensionality'.
The former refers to the general idea of reducing features from a dataset\cite{vandermaaten2007}.
In modeling this means finding the best subset of covariates which provide the most predictive power to the model or extracting the main components of the features using a \ac{PCA}.
'High-dimensionality' in contrast is a dataset attribute and applies when \(p > n\), where \(p\) is the number of covariates and \(n\) the number of observations\cite{hastie2001}.
There is no absolute value at which the term applies, both for \(n\) and \(p\).
Hence for this study the word ‘high-dimensionality‘ only applies to the
experiments using the ‘NRI‘ feature set ($>$ 7000 features for 1808 observations).

@@

Even though the feature selection task was handed over to the filter methods in this study, we ensured to not include features with a pair-wise correlation of 1.
Having such can cause undesired effects during model fitting and feature importance calculation.
To account for such, all pair-wise correlations between features were calculated.
For pairs which exceeded the threshold of 0.9999999999 the one with the largest mean absolute correlation was removed.
This process was repeated $p$ times, each time calculating a new correlation matrix.

This preprocessing step resulted in the following final number of predictors: HR (122), VI (86), NRI ().

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
\noindent Overall, the response variable 'defoliation at trees' could be modeled with an error of 28 \%.
SVM showed no differences in RMSE across feature sets whereas other learners like RF differed up to seven percentage points (HR-NRI vs. VI) (\autoref{fig:perf-result}).
Ridge and Lasso faced major issues up to the point of not achieving an error below 50 for datasets HR-NRI-VI and NRI-VI.
SVM showed the best performance of all learners with an absolute difference up around three percentage points to the next model (RF) (\autoref{tab:best-learner-perf}).
A high intra-fold variance was observed: Predicting on Luiando resulted in an RMSE value of 9.0 for learner SVM (without filter) but up to 54.26 RMSE when testing on Laukiz2 (\autoref{tab:svm-single-fold-perf}).

@@

% learner
% - SVM Car is the overall winner
% - Ridge-CV hit and miss (hit = NRI, miss = VI)
% - all MBO penalized methods have no difference across tasks
SVM combined with the 'Carscore' filter achieved the best performance (RMSE of 27.98) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with Ridge penalty (L2) showed a high variance when comparing results across tasks: In two out of six tasks (all including VI variables) the error was out of bounds (HR-NRI-VI, HR-VI, VI). For HR-NRI and HR the error was around 40 RMSE, and for task NRI the Ridge learner achieved almost the same score as the best performing model (32.9 RMSE) (\autoref{tab:perf-top-10}).
In all settings for which Ridge showed such a high error, only one observation in one fold was predicted which such a high value (in the millions).
This then caused the error estimates of these folds and the average estimate across all folds to be out of bounds.
% FIXME: All penalized methods tuned using MBO showed exactly the same error across all tasks.

@@

% latex table top 20 absolute performances
\input{performance-svm-single-plot.tex}

@@

\subsection{Variable importance}

\subsubsection{Permutation-based Variable Importance}

@@

% FIXME: explain in methods why only 2 datasets are shown
The most important features of each dataset showed an average decrease in RMSE of 1.57 (HR, B69) and 1.79 (VI, Vogelmann2) (\autoref{fig:fi-permut-vi-hr}).
For both datasets most features among the ten most important ones cluster around a wavelength range of 700 nm - 750 nm.
For feature set HR four features in the infrared region (920 nm  - 1000 nm) were identified by the model to be important due to causing a mean RMSE decrease of around one percentage point.
Overall, most features show a small importance only with average decreases in RMSE below 0.5.

@@

\noindent The decision to use a buffer of 2 m to generate the index value for each observation has to be seen critical.
Due to the reported geometric offset of up to 1 m for the hyperspectral data, the possibility of assigning a wrong pixel value to a tree observation would be high.
Using a buffer of more than 2 meters increases the probability of including information from other trees into the pixel value, blurring the actual value of the tree observation.
In our view using a buffer of 2 m is a good compromise between both worlds.
In addition, a side-analysis showed that some trees would be assigned missing values when relying on a single pixel.
This then would result in the removal of that particular tree which we wanted to avoid.
Even though we provide no results showing the influence of multiple buffer values on the extraction, we hypothesize that the relationships between features would not change substantially, leading to the almost identical model results.

Another critical point is that the exact number of contributing pixels to the final index value of an observation cannot be determined as it depends on the location of the tree within the pixel grid.
According to the extract function of the \texttt{raster} package, a pixel is included if the buffer of the point observation hits the centroid of the respective pixel, not just any part of it.
As the buffer is circular, determining the total number of contributing pixels for each tree depends on the exact location of a tree within the pixel grid.
Next, if a tree observation is located at the border of the plot, some directions of the buffer will contain no values and the subsequent index value will be calculated using less pixels than if the tree observation is located in the middle of the plot.
The magnitude of uncertainty introduced by these facts cannot be quantified.
However, we assume these to be of minor importance for the subsequent model fitting step.

\subsection{Performance vs.\ plot characteristics}
\label{subsec:perf-plot-char}

\noindent The high differences in RMSE observed for the fold performances (\autoref{tab:svm-single-fold-perf}) are a result of model overfitting.
An RMSE of 54.26 indicates a complete failure of the model in the prediction stage for this plot (Laukiz2).
Laukiz2 differs highly in the distribution of the response variable defoliation compared to all other plots (\autoref{fig:defol-distr}).
The model overfitted on learning relationships related to medium-high defoliation values but missed out on low defoliation values.
In cases when Laukiz2 is in the training set, the model is able to half the error and even reduce it down to a value of 9 RMSE in the case of Luiando.

\subsection{Predictive Performance}
The best aggregated performance of this study (SVM + Carscore filter, 27.98 RMSE) has to be seen in the light of model overfitting (see \autoref{subsec:perf-plot-char}).
Leaving out the performance on Laukiz2 when aggregating results, the mean RMSE would be 19.
However, leaving out a single plot would also change the prediction results for the other plots because the observations from Laukiz2 would not be available for model training.
With the clear presence of model overfitting in this study it can be postulated that more training data representing all health stages of a plot is needed.
A model can only make robust predictions if it has learned relationships across the whole range of the response.
Hence, care should be taken when making statements about the overall performance of the fitted models of this study due to their high variance when predicting.
However, when inspecting the fold level performances, one can say that the model does a decent job predicting defoliation ranging from 50\% to 100\% but fails for 0\% - 50\%.
This applies to all learners of this study.

\subsubsection{Algorithm differences}
An interesting find is the strength of the SVM algorithm when comparing its predictive performance to its competitors (\autoref{tab:best-learner-perf}).
% FIXME: Add refs
In the environmental community we have rarely seen a study in the last years in which SVM outperforms other models that clearly.
The absolute difference of around three percentage points is also worth noting: We could not find a study in which SVM came out as the best model in such clear way.
All of these statements are based on the assumption that a proper hyperparameter tuning was conducted for all models (which we also claim for ourselves in this study).

We were a bit surprised about the performances of the penalized methods Lasso and Ridge.
The former was not able to achieve an error below 45.96 RMSE across any dataset while Ridge came at least somewhat close to RF and XGBoost for some feature sets.
However, there were also three instances (VI, NRI-VI, HR-NRI-VI) for which the error of the fitted Ridge model was out of bounds.
We cannot preclude possible mistakes on our side when optimizing the penalization terms of the penalized learners.

% FIXME: Talk about L1 and L2 models

\subsubsection{Feature set differences}
One objective which this study aimed to answer was whether expert-based or data-driven feature engineering has a positive influence on model performance.
With respect to \autoref{fig:perf-result} one can say that overall the different feature sets did not result in substantial changes in performance.
RF and XGBoost showed a somewhat surprising decrease in performance of about six percentage points on task VI compared to all others.
We have no explanation for this specific result, especially because the SVM does not show this difference.
% FIXME: Talk about L1 and L2 models -> there we have quite some differences!

@@

% - Ensemble FS shows no advantage to simple methods -> maybe to less features, features too correlated? check diff between effect on NRI and VI
% - Effect varies across learners -> suggest to use FS even on feature sets with p < 100?
% - When calculating multiple filters, consider to only use ensemble filter?
% - PCA shows similar performance but probably requires tuning of main components -> check runtime!

@@

\subsection{Comparison to other studies}
% - no environmental studies use filter methods, some use FFS
% - only few studies analyze defoliation
% - most FS is done in bioinformatics

\noindent Most other studies analyzing defoliation operated on the plot rather than the tree level.
This is due to the low spatial resolution of used satellite products which served as the input data, making a tree-level study infeasible \cite{townsend2012, debeurs2008, rengarajan2016}.

Studies focusing on tree-level defoliation used ground-level methods such as \ac{ALS} or \ac{LiDAR}\cite{meng2018, kalin2019}.
\cite{meng2018} used \ac{OLS} regression methods while\cite{kalin2019} retrieved information out of ground-level RGB photos using \ac{CNN}.
Both study designs are substantially different compared to the setup of this work.
In addition, no spatial \ac{CV} or \ac{FS} was used.
\cite{goodbody2018} used a \ac{PLS} model with high-resolution \ac{DAP} to predict cumulative defoliation caused by the spruce budworm.
Study results indicated that spectral metrics were found to be most helpful for the model.
Incorporating such metrics (both spectral and structural) could be a possible enhancement for future works.

\cite{shendryk2016, ludwig2019} are studies which are more similar in their methodology but focus on a different response variable (woody cover).
\cite{shendryk2016} used machine learning with \ac{ALS} data to study dieback of trees for eucalyptus forests.
A grid-search was used for hyperparameter tuning and \ac{FFS} for variable selection.
\cite{ludwig2019} analyzed woody cover in South Africa using spatial \ac{CV} and \ac{FS} approach \cite{meyer2018} with a Random Forest classifier.

In summary, we could not find studies using filter methods for \ac{FS} or \ac{NRI} indices in their work with a relation to forest health.
Most studies used only one algorithm (usually Random Forest) without (strong) arguments why this particular one has been selected or why only one model was used.
This is not surprising: Most environmental/ecological datasets are not high-dimensional.
In contrast, the number of predictors if often monadic and issues related to correlations are solved manually instead of relying on an automated approach.
The bioinformatics field faces high-dimensional feature sets more often.
Hence more studies using (filter-based) feature-selection approaches can be found for this field \cite{guo2019, radovic2017}.
If a field only rarely faces high-dimensional dataset issues the motivation and expertise of using advanced methods to solve such are rather low.
We hope that this work can give some guidance and serve as a starting point for tackling high-dimensional problems in environmental modeling.

\section{Outlook and conclusion}
This study analyzed defoliation at trees in north Spain by using hyperspectral data as input for machine learning models making heavy use of filter-based feature selection methods.
It was shown that substantial differences in performance can occur depending on which feature selection methods and machine learning algorithms are combined.
SVM showed the most robust behavior across all highly-correlated datasets and was able to predict the response variable of this study substantially better than other methods.

Filter methods showed their ability to improve predictive performance for datasets with many features.
Ensemble filter methods did not show a substantial improvement over legacy filter methods.

The potential of predicting defoliation for single trees with the given toolset of this study was rather limited, seen on the average error of 27\%. 
However, with better training data covering more variety of the response variable, performance could be highly improved in future studies.
